Neutral models of \textit{de novo} gene emergence suggest that gene evolution has a preferred trajectory

### 2022-12-21 Manuscripts/deNovoRates2.tex
modified:   Modeling/DeNovoProb2.jl

\begin{equation}
P_\textit{gene-gain} = P_\textit{gene-gain}/(1-P_\textit{ORF} - P_\textit{RNA})
\label{genegaineq2}
\end{equation}

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/geneGainLossComplete_new.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}. We note that \autoref{genegaineq} describes total gene gain probability.}
\label{gainlossprob}
\vspace{1ex}
\hrule
\end{figure}

@@

For example, a log-log ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event. We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF. We found that for ORF longer than 44 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}). To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for a proto-gene containing an ORF with 50 codons, at different values of GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}). For this proto-gene, we found that the log-log ratio initially decreases with GC-content. It is minimum at a GC-content of 0.41, above which it increases. More generally, the ratio remains above 2 for all GC-content values for this proto-gene.

@@

\begin{equation}
P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain}
\label{orffirsteq1}
\end{equation}

@@

\begin{equation}
P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain}
\label{rnafirsteq1}
\end{equation}

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/whichfirstComplete_with2step.pdf}
\caption{Proto-genes with shorter ORFs preferentially emerge from existing ORFs (ORF-first) whereas proto-genes with longer ORFs preferentially emerge from existing RNAs (RNA-first). \textbf{(A)} The horizontal axis denotes the number of codons in the ORF. The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence, with a GC-content of 0.42 (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}). A positive value suggests an ORF-first trajectory and a negative value supports an RNA-first trajectory. Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}). \textbf{(B)} Minimum ORF length (codons) that makes RNA-first trajectory more feasible (vertical axis) for different GC-content (horizontal axis). For a GC-content of 0.3 and 0.35, proto-genes of every analysed ORF length ($\geq33$ codons) preferentially emerge RNA-first.}
\label{whoisfirst}
\vspace{1ex}
\hrule
\end{figure}
We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first). In other words, the ORF-first hypothesis is more feasible. Likewise, a negative value of the ratio would suggest that the RNA-first hypothesis is more feasible. With the specific parameter values we used, it turns out that proto-genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}). This minimum ORF length above which the RNA-first hypothesis is more feasible than ORF-first hypothesis, increases with higher GC-content. For example, this number is 48 codons for a GC-content of 0.45 (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).

>>>>

\begin{equation}
P_\textit{gene-gain} = P_\textit{gene-gain}^*/(1-P_\textit{ORF} - P_\textit{RNA})
\label{genegaineq2}
\end{equation}

@@

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/geneGainLossComplete2_new.pdf}
\caption{Genes are more likely to be independently lost twice, than being born once. The vertical axis in both panels shows the number of independent gene loss events (\autoref{genelosseq}) that can occur relative to one gene gain event (\autoref{genegaineq2}) -- $\log(P_\textit{gene-gain})/\log(P_\textit{gene-loss})$. The horizontal axis shows the number of codons in the ORF \textbf{(A)}, and the GC-content \textbf{(B)}.}
\label{gainlossprob}
\vspace{1ex}
\hrule
\end{figure}

@@

For example, a log-log ratio value of 2 would indicate that two independent gene loss events are as likely as a single gene gain event. We calculated this ratio for genes with different ORF lengths, because ORF gain and loss probabilities depend on the length of the ORF. We found that for ORF longer than 44 codons, at least two independent gene loss events can occur in the time frame of one gene gain event (\hyperref[gainlossprob]{\autoref{gainlossprob}A}). To understand how GC-content determines the likelihood of gene gain relative to gene loss, we calculated the log-log ratio of the two probability values for proto-genes with different GC-content (\hyperref[gainlossprob]{\autoref{gainlossprob}B}), and that contain ORFs of different lengths (codons). We found that the effect of GC-content on the log-log ratio depends on the length of the ORF. Specifically, the log-log ratio steadily increases with GC-content ranging from 0.3 to 0.6, when the proto-gene contains an ORF with 30 codons. When the ORF has 60 -- 90 codons, the log-log ratio initially decreases with GC content and then increases. For ORFs containing more than 120 codons, the log-log ratio decreases with increasing GC-content (within the analysed range). Overall, beyond a certain ORF length, gene loss could occur at least two times independently in the time frame of one gene gain event, irrespective of the GC content.

@@

\begin{equation}
P_\textit{ORF-first} = P_\textit{ORF-stay}\times P_\textit{RNA-gain \vline\ ORF}
\label{orffirsteq1}
\end{equation}

@@

\begin{equation}
P_\textit{RNA-first} = P_\textit{RNA-stay}\times P_\textit{ORF-gain \vline\ RNA}
\label{rnafirsteq1}
\end{equation}

\begin{figure}[!t]
\centering
\includegraphics[scale=0.5]{Figures/whichfirstComplete2_new.pdf}
\caption{Proto-genes with shorter ORFs preferentially emerge from existing ORFs (ORF-first) whereas proto-genes with longer ORFs preferentially emerge from existing RNAs (RNA-first). \textbf{(A)} The horizontal axis denotes the number of codons in the ORF. The vertical axis shows the log transformed ratio of single-step probabilities of ORF-first and RNA-first trajectories of \textit{de novo} emergence, with a GC-content of 0.42 (Equations \ref{orffirsteq1} \& \ref{rnafirsteq1}). A positive value suggests an ORF-first trajectory and a negative value supports an RNA-first trajectory. Similar analysis with two-step probabilities shows a smaller difference between the likelihood of the two trajectories (Equations \ref{orffirsteq2} \& \ref{rnafirsteq2}). \textbf{(B)} Minimum ORF length (codons) that makes RNA-first trajectory more feasible (vertical axis) for different GC-content (horizontal axis). For a GC-content of 0.3 and 0.35, proto-genes of every analysed ORF length ($\geq33$ codons) preferentially emerge RNA-first.}
\label{whoisfirst}
\vspace{1ex}
\hrule
\end{figure}
We calculated the log-transformed ratio of $P_\textit{ORF-first}$ and $P_\textit{RNA-first}$, such that a positive value of the ratio would mean that transcription gain for an untranscribed ORF (ORF-first) is more likely than ORF gain in an already transcribed DNA (RNA-first). In other words, the ORF-first hypothesis is more feasible. Likewise, a negative value of the ratio would suggest that the RNA-first hypothesis is more feasible. With the specific parameter values we used, it turns out that proto-genes with up to 42 codons in their coding region, preferentially emerge ORF-first (\hyperref[whoisfirst]{\autoref{whoisfirst}A}). This minimum ORF length above which the RNA-first hypothesis is more feasible than ORF-first hypothesis, increases with higher GC-content. For example, this number is 48 codons for a GC-content of 0.45 (\hyperref[whoisfirst]{\autoref{whoisfirst}B}).
